Type cones of $\b{g}$-vector fans

### 2019-02-01 thirdPaper/draft/typeConeAssociahedra.tex
More on Section 3

% PG et Yann
\newcommand{\field}{\mathbb{K}}
\newcommand{\cat}{\mathcal{C}}
\newcommand{\Hom}[1]{\operatorname{Hom}_{#1}}
\newcommand{\susp}{\Sigma}
\newcommand{\add}{\operatorname{add}}
\newcommand{\MOD}{\operatorname{mod}}
\newcommand{\End}[1]{\operatorname{End}_{#1}}
\newcommand{\spl}{\operatorname{sp}}
\newcommand{\Ksp}{K_0^{\spl}}
\newcommand{\Kg}{K_0^{\b{g}}}
\newcommand{\ind}{\operatorname{ind}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Relations for~$\b{g}$-vectors in~$2$-Calabi--Yau triangulated categories}
Let~$\field$ be a field.  Let~$\cat$ be a~$\field$-linear triangulated category with suspension functor~$\susp$.  We fix a collection~$\ind(\cat)$ of representatives of isomorphism classes of indecomposable objects of~$\cat$. We will assume the following:
\begin{itemize}
 \item $\cat$ is essentially small (in particular,~$\ind(\cat)$ is a set);
 \item $\cat$ is~$\Hom{}$-finite: for each pair of objects~$X$ and~$Y$, the~$\field$-vector space~$\cat(X,Y)$ is finite-dimensional;
 \item $\cat$ is Krull--Schmidt: the endomorphism algebra of any indecomposable object is local;
 \item $\cat$ is~$2$-Calabi--Yau: for each pair of objects~$X$ and~$Y$, there is an isomorphism of bifunctors
 \[
  \cat(X,\susp Y) \to D\cat(Y,\susp X)
 \]
 where~$D=\Hom{\field}(-,\field)$ is the usual duality of vector spaces;
 \item $\cat$ contains a basic cluster-tilting object~$T=\bigoplus_{i=1}^n T_i$:
 \[
  \textrm{for any object~$X$,} \quad \cat(T, \susp X) = 0 \quad \textrm{if and only if} \quad X\in \add(T),
 \]
 where~$\add(T)$ is the smallest additive full subcategory of~$\cat$ containing the~$T_i$'s and closed under isomorphisms.
\end{itemize}

We need to introduce some notations before stating the main result of this section.  Let~$\Lambda:= \End{\cat}(T)$, and let~$F$ be the functor
\[
 F=\cat(T,-):\cat \xrightarrow{} \MOD \Lambda.
\]
\begin{proposition}[]\pg{Add references to BMR and KR}
 The functor~$F$ induces an equivalence of~$\field$-linear categories
 \[
  F:\cat/(\susp T) \xrightarrow{} \MOD \Lambda,
 \]
 where~$(\susp T)$ is the ideal of morphisms factoring through an object of~$\add(\susp T)$ and~$\MOD \Lambda$ is the category of finite-dimensional right~$\Lambda$-modules.  This equivalence induces further equivalences between~$\add(T)$ and the category of projective modules, and between~$\add(\susp^2 T)$ and that of injective modules.
\end{proposition}

@@

 \item For any indecomposable object~$Y$ of~$\cat$, let
 \[
  \susp Y \to E' \to  Y \to \susp^2 Y
 \]
 be an almost split triangle (unique up to isomorphism).  We let
 \[
  r_Y := [Y] + [\susp Y]  - [E'] \in \Ksp(\cat).
 \]
\end{enumerate}

@@

\begin{theorem}
 Let~$\cat$ be a category satisying the hypotheses of \ref{} \pg{add ref}.  Then~$\cat$ has only finitely many isomorphism classes of indecomposable objects if and only if the set~\[L:=\big\{\ell_X \ | \ X\in \ind(\cat)\setminus \add(\susp T) \big\}\] generates the kernel of~$\b g:\Ksp(\cat) \to \Kg(\cat)$. 
 In this case, the set~$L$ is a basis of the kernel of~$\b g$, and \pg{add formula}.
\end{theorem}

\end{notation}

>>>>

% PG et Yann
\newcommand{\field}{\mathbb{K}}
\newcommand{\cat}{\mathcal{C}}
\newcommand{\Hom}[1]{\operatorname{Hom}_{#1}}
\newcommand{\susp}{\Sigma}
\newcommand{\add}{\operatorname{add}}
\newcommand{\MOD}{\operatorname{mod}}
\newcommand{\End}[1]{\operatorname{End}_{#1}}
\newcommand{\spl}{\operatorname{sp}}
\newcommand{\Ksp}{K_0^{\spl}}
\newcommand{\Kg}{K_0^{\b{g}}}
\newcommand{\ind}{\operatorname{ind}}
\newcommand{\coker}{\operatorname{coker}}
\newcommand{\CC}{\phi}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Relations for~$\b{g}$-vectors in~$2$-Calabi--Yau triangulated categories}
%%%%%
\subsection{Setting}
Let~$\field$ be a field.  Let~$\cat$ be a~$\field$-linear triangulated category with suspension functor~$\susp$.  We fix a collection~$\ind(\cat)$ of representatives of isomorphism classes of indecomposable objects of~$\cat$. We will assume the following:
\begin{itemize}
 \item $\cat$ is essentially small (in particular,~$\ind(\cat)$ is a set);
 \item $\cat$ is~$\Hom{}$-finite: for each pair of objects~$X$ and~$Y$, the~$\field$-vector space~$\cat(X,Y)$ is finite-dimensional;
 \item $\cat$ is Krull--Schmidt: the endomorphism algebra of any indecomposable object is local;
 \item $\cat$ is~$2$-Calabi--Yau: for each pair of objects~$X$ and~$Y$, there is an isomorphism of bifunctors
 \[
  \cat(X,\susp Y) \to D\cat(Y,\susp X)
 \]
 where~$D=\Hom{\field}(-,\field)$ is the usual duality of vector spaces;
 \item $\cat$ contains a basic cluster-tilting object~$T=\bigoplus_{i=1}^n T_i$:
 \[
  \textrm{for any object~$X$,} \quad \cat(T, \susp X) = 0 \quad \textrm{if and only if} \quad X\in \add(T),
 \]
 where~$\add(T)$ is the smallest additive full subcategory of~$\cat$ containing the~$T_i$'s and closed under isomorphisms.
\end{itemize}

%%%%
\subsection{Statement of the theorem}
We need to introduce some notations before stating the main result of this section.  Let~$\Lambda:= \End{\cat}(T)$, and let~$F$ be the functor
\[
 F=\cat(T,-):\cat \xrightarrow{} \MOD \Lambda.
\]
\begin{proposition}[]\pg{Add references to BMR and KR}
 The functor~$F$ induces an equivalence of~$\field$-linear categories
 \[
  F:\cat/(\susp T) \xrightarrow{} \MOD \Lambda,
 \]
 where~$(\susp T)$ is the ideal of morphisms factoring through an object of~$\add(\susp T)$ and~$\MOD \Lambda$ is the category of finite-dimensional right~$\Lambda$-modules.  This equivalence induces further equivalences between~$\add(T)$ and the category of projective modules, and between~$\add(\susp^2 T)$ and that of injective modules.
\end{proposition}

@@

 \item For any indecomposable object~$Y$ of~$\cat$, let
 \[
  \susp Y \to E' \to  Y \to \susp^2 Y
 \]
 be an almost split triangle (unique up to isomorphism).  We let
 \[
  r_Y := [Y] + [\susp Y]  - [E'] \in \Ksp(\cat).
 \]
\end{enumerate}
\end{notation}

@@

\begin{theorem}\label{theo::relations-g-vecteurs}
 Let~$\cat$ be a category satisying the hypotheses of \ref{} \pg{add ref}.  Then~$\cat$ has only finitely many isomorphism classes of indecomposable objects if and only if the set~\[L:=\big\{\ell_X \ | \ X\in \ind(\cat)\setminus \add(\susp T) \big\}\] generates the kernel of~$\b g:\Ksp(\cat) \to \Kg(\cat)$. 
 In this case, the set~$L$ is a basis of the kernel of~$\b g$, and \pg{add formula}, for any $x\in \ker(\b g)$, we have that
 \[
  x= \sum_{A\in \ind(\cat) \setminus \add(\susp T)} \frac{\langle x, A \rangle}{\langle \ell_A, A \rangle} \ell_A.
 \]
\end{theorem}

\begin{corollary}
 Assume that~$\ind(\cat)$ is finite.  Let~$X\xrightarrow{} E \xrightarrow{} Y \xrightarrow{h} \susp X$ be a triangle with~$h\in (\susp T)$.  Then the element~$x=[X]+[Y]-[E]$ of the kernel of~$\b g$ is a non-negative linear combination of the~$\ell_A$, with~$A\in \ind(\cat)\setminus \add(\susp T)$.
\end{corollary}
\begin{proof}
 We know that~$x=\sum_{A\in \ind(\cat)\setminus \add(\susp T)} \frac{\langle x, [A] \rangle}{\langle \ell_A, [A] \rangle} \ell_A$; since~$\langle \ell_A, [A] \rangle$ is positive by Lemma~\ref{lemm::bilinear-form}, we only need to show that each~$\langle x, [A] \rangle$ is non-negative. 
 The functor~$F=\cat(T,-)$ induces an exact sequence
 \[
  FX \to FE \to FY \to 0,
 \]
 which in turn induces an exact sequence
 \[
  0\to \Hom{\Lambda}(FY, FA) \to \Hom{\Lambda}(FE, FA) \xrightarrow{f} \Hom{\Lambda}(FX, FA) \to \coker(f) \to 0.
 \]
 Therefore~$\langle x, [A] \rangle = \dim_{\field} \coker(f) \geq 0$.
\end{proof}

%%%%
\subsection{Proof of Theorem \ref{theo::relations-g-vecteurs}}

\begin{lemma}
 If~$X$ or~$Y$ lie in~$\add(\susp T)$, then~$\langle X, Y \rangle = 0$.
\end{lemma}
\begin{proof}
 This is because~$F\susp T = 0$.
\end{proof}

\begin{lemma}\label{lemm::almost-split}
 Let~$X\xrightarrow{} Y \xrightarrow{} \susp^{-1} X \xrightarrow{h} \susp X$ be an almost-split triangle.  Then~$X\notin \add(\susp T)$ if and only if~$h\in (\susp T)$.
\end{lemma}
\begin{proof}
 If~$X\in \add(\susp T)$, then~$h$ cannot be in~$(\susp T)$, otherwise it would be zero since~$\cat(T, \susp T) =0$.

 Assume now that~$X\notin \add(\susp T)$.  Let~$\field_X$ be the residue field of the algebra~$\End{\cat}(X)$. 
 By definition of an almost-split triangle,~$h$ is in the socle of the right~$\End{\cat}(X)$-module~$\cat(\susp^{-1} X, \susp X)$.  Moreover, this socle is a one-dimensional~$\field_X$-vector space; indeed, the~$2$-Calabi--Yau condition gives an isomorphism
 \[
  \cat(\susp^{-1} X, \susp X) \cong D\cat(X, X).
 \]
 Thus the socle of the right module~$\cat(\susp^{-1} X, \susp X)$ has the same~$\field_X$ dimension as the top of the left module~$\cat(X,X)$.  Since~$X$ is indecomposable,~$\cat(X,X)$ is local, and its top is one-dimensional over~$\field_X$.

 Now,~$(\susp T)\cat(\susp^{-1} X, \susp X)$ is a sub-module of~$\cat(\susp^{-1} X, \susp X)$.  Therefore, if~$(\susp T)\cat(\susp^{-1} X, \susp X)$ is non-zero, then it contains the one-dimensional socle of~$\cat(\susp^{-1} X, \susp X)$, and thus contains~$h$.   
 By \cite{} \pg{cite Palu},
 \[
  (\susp T)\cat(\susp^{-1} X, \susp X) \cong D\cat(X, X)/(\susp T). 
 \]
The identity morphism of~$X$ is not in~$(\susp T)$, since~$X$ is not in~$\add(\susp T)$.  Thus the right-hand side is non-zero, and so neither is the left-hand side.  By the above, this implies that~$h\in (\susp T)\cat(\susp^{-1} X, \susp X)$, which finishes the proof.
\end{proof}

\begin{lemma}\label{lemm::bilinear-form}
Let~$A$ and~$B$ be two indecomposable objects of~$\cat$. \pg{introduce general notation for redisue field}
\begin{enumerate}
 \item If~$A \notin \add(\susp T)$, then
 \[
  \langle \ell_A, B \rangle = \begin{cases}
                               0 & \textrm{if~$A\ncong B$;} \\
                               \dim_{\field}\field_A & \textrm{if~$A\cong B$.}
                              \end{cases}
 \]

 \item If~$B \notin \add(\susp T)$, then
 \[
  \langle A, r_B \rangle = \begin{cases}
                               0 & \textrm{if~$A\ncong B$;} \\
                               \dim_{\field}\field_B & \textrm{if~$A\cong B$.}
                              \end{cases}
 \]
\end{enumerate} 
\end{lemma}
\begin{proof}
 We only prove the first assertion; the second one is proved dually.  Assume that~$A\notin\add(\susp T)$.  Let
 \[
  A \xrightarrow{f} E \xrightarrow{} \susp^{-1} A \xrightarrow{h} \susp A
 \]
 be an almost-split triangle.  By Lemma~\ref{lemm::almost-split}, the morphism~$h$ is in~$(\susp T)$.  Applying the functor~$F=\cat(T,-)$, we get an exact sequence
 \[
  FA\xrightarrow{Ff} FE \xrightarrow{} F\susp^{-1} A \xrightarrow{} 0.
 \]
 Applying now the functor~$\Hom{\Lambda}(-, FB)$, we get an exact sequence
 \[
  0 \xrightarrow{} \Hom{\Lambda}(F\susp^{-1} A, FB) \xrightarrow{} \Hom{\Lambda}(FE, FB) \xrightarrow{Ff^*} \Hom{\Lambda}(FA, FB) \xrightarrow{} \coker(Ff^*) \xrightarrow{} 0
 \]
 If~$B\ncong A$, then the definition of an almost-split triangle implies that~$Ff^*$ is surjective. \pg{add an exact sequence here; image of Ffstar is non-invertible elements} Thus~$\coker(Ff^*) = 0$, and by additivity of the dimension in exact sequences, we get that~$\langle \ell_A, B \rangle =0$.

 If~$B\cong A$, then~$\coker(Ff^*)$ is isomorphic to the residue field of~$\End{\Lambda}(FA)$, which is isomorphic to the residue field~$\field_A$ of~$A$.
\end{proof}

\pg{TO DO}

%%%%
\subsection{Application to~$\b g$-vectors}

We recall the following results and definition from \cite{} \pg{cite DK and Palu}.

\begin{proposition}
 Let~$X$ be an object of~$\cat$. Then there exists a triangle
 \[
  T_1^X \to T^0_X \to X \to \susp T_1^X
 \]
 with~$T_1^X$ and~$T_0^X$ in~$\add(T)$.

\end{proposition}

\begin{definition}
 The \defn{index} of an object~$X$ is the element
 \[
  \ind_T(X) := [T_0^X] - [T_1^X] \in \Ksp(\add(T)).
 \]
\end{definition}

The notion of index is very close to the definition of the map~$\b g:\Ksp(\cat)\to \Kg(\cat)$ of Definition~\ref{} \pg{ref}.  The link is given by the following result.

\begin{proposition}[]\pg{cite Palu}
 Let~$X\xrightarrow{} Y \xrightarrow{} Z \xrightarrow{h} \susp X$ be a triangle.
 Then
 \[
  \ind_T(X) + \ind_T(Z) - \ind_T(Y) = 0
 \]
 if and only if~$h\in (\susp T)$.
\end{proposition}

\begin{corollary}
 There is an isomorphism~$\phi:\Kg(\cat) \to \Ksp(\add(T))$ such that~$\ind_T = \phi\circ \b g$.  In particular,~$\Kg(\cat)$ is a free abelian group generated by the~$[T_i]$.
\end{corollary}

\begin{theorem}[]\pg{cite???}
 Let~$\cat$ be the cluster category of a valued quiver of Dynkin type~$A,B,C,D,E,F$ or~$G$.  Let~$T$ be a basic cluster-tilting object in~$\cat$, and let~$Q$ be the valued Gabriel quiver of~$\End{\cat}(T)$.  Then there is a bijection
 \[
  \CC: \ind(\cat) \xrightarrow{} \{\textrm{cluster variables in the cluster algebra of~$Q$}\},
 \]
 which has the follwing properties.
 \begin{itemize}
  \item For any~$X,Y\in \ind(\cat)$, $\CC(X)$ and~$\CC(Y)$ are compatible if and only if~$\cat(X, \susp Y) = 0$.
  \item For any~$X\in \ind(\cat)$, the~$\b g$-vector of~$\CC(X)$ is~$\b g([X])$, where we identify~$\Z^n$ and~$\Kg(\cat)$ via the isomorphism sending~$(a_1, \ldots, a_n)$ to~$\sum_{i=1}^n a_i [T_i]$.
  \item For any~$X,Y\in \ind(\cat)$, $\CC(X)$ and~$\CC(Y)$ are exchangeable if and only if~$\dim_{\field_X}\cat(X, \susp Y) = \dim_{\field_Y}\cat(X, \susp Y) = 1$.
 \end{itemize}
\end{theorem}

The cluster category also contains enough information to detect positive mutations.

\begin{proposition}
 Let~$X,Y\in \ind(\cat)$ be such that~$\dim_{\field_X}\cat(X, \susp Y) = \dim_{\field_Y}\cat(X, \susp Y) = 1$.  Let 
 \[
  X\xrightarrow{} E \xrightarrow{} Y \xrightarrow{h} \susp X \quad \textrm{and} \quad Y\xrightarrow{} E' \xrightarrow{} X \xrightarrow{h'} \susp Y
 \]
 be triangles with~$h$ and~$h'$ non-zero (these triangles are unique up to isomorphism).  Then~$\CC(Y)$ is obtained by performing a positive mutation on~$\CC(X)$ in some seed if and only if~$h\in (\susp T)$.
 \begin{proof}
  \pg{TO DO, Vincent: definition of positive mutation?}
 \end{proof}
\end{proposition}
